---
title: "UKB_SPQ"
author: "Yuanjun Gu"
date: "2026-01-16"
output: html_document
---

```{r setup, include=FALSE}
# Check path variables auto load in .Rprofile
library(here) # for relative path finding
library(tidyverse)
library(data.table)
library(glue)

# for general visualisation
library(patchwork)

# for phenotype correlation
library(Hmisc)
library(corrplot)

# for GWAS visualisation
library(topr)
```
# Generate Master Pheno/Covar Table for SPQ and calculate SPQ_total

Sensory Perception Quotient-10 (SPQ-10)
For further description please go to: https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=520

For all other traits/phenotypes/etc, check respective entry through these files:
~/rds/rds-rb643-ukbiobank2/Data_Phenotype/DataFetch_09052025/ukb679961.html - All fields
~/rds/rds-rb643-ukbiobank2/Data_Phenotype/DataFetch_22122025/ukb704872.html - Online follow-up Questionnaires

If you still didn't find it, then you should be able to find them all at here: https://biobank.ndph.ox.ac.uk/showcase/search.cgi

```{r Dataframe formatting and Cleaning: Initial Listed Columns of Interests}
# Grabbing potentially related items of interests
# eid - UKB annonymised id

# age-related items:
#   Data-Field 52: Month of birth
#   Data-Field 34: Year of birth
#   32362	When online Sensory Perception Quotient-10 (SPQ-10) module completed

# sex/gender-related items:
#   31-0.0 Reported Sex
#   22001-0.0 Genetic Sex

# SPQ10 items:
#   32353	Ability to distinguish between apparently identical items of clothing
#   32345	Ability to distinguish different people by their smell
#   32348	Ability to notice a tiny change in the temperature
#   32354	Ability to see dust particles in the air
#   32351	Ability to smell a small gas leak within the house
#   32349	Ability to taste the difference between apparently identical sweets
#   32346	Ability to taste the difference between two brands of salty potato crisps
#   32350	Ability to tell the weight difference between two coins with eyes closed
#   32347	Can hear electricity humming in the walls
#   32352	Would be the first to hear a fly in the room

# Physical Health related items: [!!! May have more appropriate item list, do double check !!!]
# Data-Field 29087: Experienced a life-threatening injury or illness
# Data-Field 29201	When online Adverse life events questionnaire completed
# https://biobank.ndph.ox.ac.uk/showcase/search.cgi?wot=0&srch=diagnos&sim=on&yfirst=2000&ylast=2026 = This probably covers everything you ever need about this.

# Mental Health related items:
# Data-Field 29000: Using this in particular as it is newer, includes more people, more conditions and allow multiple selections
# Alternatively you can use: Data-Field 20544, but this was done in 2016

# ADHD specific:
# Data-Field 32262 Diagnosis of ADHD by a medical professional = Data-Field 32262(N=209) vs Data-Field 29000 (N=241), maybe think about what to chose.
# Data-Field 32263 Age of ADHD diagnosis

# ASD specific:
# Data-Field 32324 Diagnosis of autism by a medical professional = Data-Field 32324(N=313) vs Data-Field 29000 (N=408), maybe think about what to chose.
# Data-Field 32325 Age of autism diagnosis 

# GWAS QC-related items:
# 21000-0.0 Self-identified white Europeans
# 22027-0.0 Outliers for heterozygosity or missing rate
# 22000-0.0 genotype batch
# genetic PCA": "22009-0.1","22009-0.2","22009-0.3","22009-0.4","22009-0.5","22009-0.6","22009-0.7","22009-0.8","22009-0.9","22009-0.10",
#                "22009-0.11","22009-0.12","22009-0.13","22009-0.14","22009-0.15","22009-0.16","22009-0.17","22009-0.18","22009-0.19","22009-0.20"
#                "22009-0.21", "22009-0.22", "22009-0.23","22009-0.24", "22009-0.25", "22009-0.26", "22009-0.27", "22009-0.28", "22009-0.29","22009-0.30",
#                "22009-0.31", "22009-0.32", "22009-0.33", "22009-0.34", "22009-0.35","22009-0.36", "22009-0.37", "22009-0.38", "22009-0.39", "22009-0.40"
```

```{r Dataframe formatting and Cleaning: Format Columns of Interests List to use for search}
# Participant ID
eid_vars = "eid"

# Age-related variables
age_vars = c(
  "52-0.0",      # Month of birth
  "34-0.0",      # Year of birth
  "32362-0.0"    # When online SPQ-10 module completed
)

# Sex/gender-related variables
sex_vars = c(
  "31-0.0",    # Reported Sex
  "22001-0.0"  # Genetic Sex
)

# SPQ-10 (Sensory Perception Quotient) items
spq10_vars = c(
  "32353-0.0",  # Ability to distinguish between apparently identical items of clothing
  "32345-0.0",  # Ability to distinguish different people by their smell
  "32348-0.0",  # Ability to notice a tiny change in the temperature
  "32354-0.0",  # Ability to see dust particles in the air
  "32351-0.0",  # Ability to smell a small gas leak within the house
  "32349-0.0",  # Ability to taste the difference between apparently identical sweets
  "32346-0.0",  # Ability to taste the difference between two brands of salty potato crisps
  "32350-0.0",  # Ability to tell the weight difference between two coins with eyes closed
  "32347-0.0",  # Can hear electricity humming in the walls
  "32352-0.0"   # Would be the first to hear a fly in the room
)

# # !! TODO: DECIDE WHAT TO DO FOR IF NEEDED !! Physical health variables
# physical_health_vars = c(
#   "29087",  # Experienced a life-threatening injury or illness
#   "29201"   # When online Adverse life events questionnaire completed
# )
# # Note: For diagnoses, see: https://biobank.ndph.ox.ac.uk/showcase/search.cgi?wot=0&srch=diagnos&sim=on&yfirst=2000&ylast=2026

# Mental health variables
mental_health_vars = paste0("29000-0.", 0:19) # Multiple choice mental health diagnoses (newer, more comprehensive)
# Alternative: "20544" (older, from 2016)

# ADHD-specific variables
adhd_vars = c(
  "32262-0.0",  # Diagnosis of ADHD by a medical professional (N=209 vs 29000 N=241)
  "32263-0.0"   # Age of ADHD diagnosis
)

# ASD-specific variables
asd_vars = c(
  "32324-0.0",  # Diagnosis of autism by a medical professional (N=313 vs 29000 N=408)
  "32325-0.0"   # Age of autism diagnosis
)

# GWAS QC variables
gwas_qc_vars = c(
  "22027-0.0",  # Outliers for heterozygosity or missing rate
  "22000-0.0"   # Genotype batch
)

# Genetic PCs (first 40 principal components)
genetic_pca_vars = paste0("22009-0.", 1:40)

# Combine all variables
all_vars = c(
  eid_vars,
  age_vars,
  sex_vars,
  spq10_vars,
  mental_health_vars,
  adhd_vars,
  asd_vars,
  gwas_qc_vars,
  genetic_pca_vars
)

```

```{r Dataframe formatting and Cleaning: Compiling the master table, eval=FALSE, include=FALSE}
# Select through fread to minimise reading time and cost
UKB.22122025 = fread("~/rds/rds-rb643-ukbiobank2/Data_Phenotype/DataFetch_22122025/ukb704872.csv", select =all_vars) 
# This table includes only mental health info
UKB.09052025 = fread("~/rds/rds-rb643-ukbiobank2/Data_Phenotype/DataFetch_09052025/ukb679961.csv", select=all_vars) 
# includes everything except "21000-0.0"

# # omit people who did not finish SPQ
UKB.22122025.f = UKB.22122025  %>% drop_na(any_of(spq10_vars))
 
# # merge so we have important data for everyone had SPQ data
UKB.f <- UKB.22122025.f %>% left_join(UKB.09052025, by = "eid", suffix = c("", ".old")) 
# in case there are duplicated entries other than eid

write.csv(UKB.f, here("UKB_master_pheno_table.txt"), row.names = FALSE)
```

```{r Sort SPQ coding, eval=FALSE, include=FALSE}
# According to Data Code Dictionary for SPQ-10:
# Left = Response               Right = Translated SPQ Score
# 1	Strongly agree            = 3
# 2	Agree                     = 2
# 3	Disagree                  = 1
# 4	Strongly disagree         = 0
# -3	Prefer not to answer    = NA

UKB.f = fread(here("UKB_master_pheno_table.txt"))

UKB.f.recoded = UKB.f %>% 
  mutate(across(all_of(spq10_vars), 
                ~recode(., `1` = 3, `2` = 2, `3` = 1, `4` = 0, `-3` = NA_real_)))

# No imputation for SPQ due to small questionnaire size

UKB.f.recoded.naomit = UKB.f.recoded %>% drop_na(any_of(spq10_vars)) # 149,961 people left
UKB.f.recoded.naomit.withtotalSPQ = UKB.f.recoded.naomit %>% 
  mutate(SPQ_total = rowSums(across(all_of(spq10_vars))))

write.csv(UKB.f.recoded.naomit.withtotalSPQ, here("UKB_master_pheno_table_SPQ.recoded.summed.txt"), row.names = FALSE)
```

# Participants QC - generate keep.id.list
1. Remove Excessive heterozygosity {keep 22027-0.0 == NA}
2. Find Reported Sex and Genetic Sex Mismatch {keep 31-0.0 == 22001-0.0, ignore sex in NA if genetic sex is present}
3. Remove gPCA outliers using PCA1 and PCA2 to ensure genetically homogeneous white European population 
   {keep "22009-0.1" within its mean+-5SD and "22009-0.2" within its mean+-5SD}

```{r Participants QC code, eval=FALSE, include=FALSE}
# Step 1. remove excessive heterozygosity 
UKB.f.withtotalSPQ = fread(here("UKB_master_pheno_table_SPQ.recoded.summed.txt"))
UKB.f.withtotalSPQ.low_hetero = filter(UKB.f.withtotalSPQ, is.na(UKB.f.withtotalSPQ$`22027-0.0`))

# Step 2. exclude mismatch sex
UKB.f.withtotalSPQ.low_hetero.withsexQC <- UKB.f.withtotalSPQ.low_hetero %>%
  filter(
    !is.na(`22001-0.0`) &                           # Must have genetic sex
    (is.na(`31-0.0`) | `31-0.0` == `22001-0.0`)     # Either missing self-report OR it matches
  )

# Step 3. Remove gPCA outliers to ensure population is genetically homogeneous
PC1_mean <- mean(UKB.f.withtotalSPQ.low_hetero.withsexQC$`22009-0.1`, na.rm = TRUE)
PC1_sd   <- sd(UKB.f.withtotalSPQ.low_hetero.withsexQC$`22009-0.1`, na.rm = TRUE)
PC2_mean <- mean(UKB.f.withtotalSPQ.low_hetero.withsexQC$`22009-0.2`, na.rm = TRUE)
PC2_sd   <- sd(UKB.f.withtotalSPQ.low_hetero.withsexQC$`22009-0.2`, na.rm = TRUE)

UKB.f.withtotalSPQ.low_hetero.withsexQC.homogeneous <- UKB.f.withtotalSPQ.low_hetero.withsexQC %>%
  filter(
    between(`22009-0.1`, PC1_mean - 5*PC1_sd, PC1_mean + 5*PC1_sd),
    between(`22009-0.2`, PC2_mean - 5*PC2_sd, PC2_mean + 5*PC2_sd)
  )

write.csv(UKB.f.withtotalSPQ.low_hetero.withsexQC.homogeneous, file= here("UKB_master_pheno_withSPQtotal.QCed.txt"), row.names = FALSE)
```

```{r Generate formatted master phenotype table by filtering out none-AQ people and generating FID, eval=FALSE}
AQ.keep.id = fread("/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/AQ/UKBAQ_keep.id.QCed.txt")
SPQ.id = fread(here("UKB_master_pheno_withSPQtotal.QCed.txt"))

sum(!SPQ.id$eid %in% AQ.keep.id$f.eid) # 2754 SPQ population not in AQ population 

# Note. These population were removed so GRM for AQ can be applied for the GWAS to keep SPQ and AQ complementary to each other.

SPQ.keep.id = SPQ.id[SPQ.id$eid %in% AQ.keep.id$f.eid, ] %>%
  mutate(FID = eid, IID = eid) %>%
  mutate(
    dob = as.Date(paste0(`34-0.0`, "-", `52-0.0`, "-01")),
    age_years = round(interval(dob, `32362-0.0`) / months(1) / 12)
  ) %>%
  mutate(SPQ_total_scaled = as.numeric(scale(SPQ_total))) %>%
  select(FID, IID, age_years, everything(), -eid, -`52-0.0`, -`34-0.0`, -`32362-0.0`, -dob)
# Format and generate FID, IID and age so it can be easy sliced into mpheno, COVAR and qCOVAR for GCTA-fastGWA to run.
# Create SPQ_total_scaled for good practice.

write.table(SPQ.keep.id, file = here("UKB_master_pheno_withSPQtotal.QCed.AQfiltered.formatted.txt"), row.names = FALSE, col.names = TRUE)
# write final formatted results for later splitting.
```

# GWAS: mpheno, COVAR and qCOVAR generation
```{r mpheno to loop through in GWAS, eval=FALSE, include=FALSE}
spq10_labels = c(
  "32345-0.0" = "SPQ_smell_people",
  "32346-0.0" = "SPQ_taste_crisps",
  "32347-0.0" = "SPQ_hearing_electricity",
  "32348-0.0" = "SPQ_touch_temperature",
  "32349-0.0" = "SPQ_taste_sweets",
  "32350-0.0" = "SPQ_touch_coinweight",
  "32351-0.0" = "SPQ_smell_gas",
  "32352-0.0" = "SPQ_hearing_fly",
  "32353-0.0" = "SPQ_vision_clothing",
  "32354-0.0" = "SPQ_vision_dust"
)

SPQ_master = fread(here("UKB_master_pheno_withSPQtotal.QCed.AQfiltered.formatted.txt"))
SPQ_pheno = SPQ_master %>%
  rename(!!!setNames(names(spq10_labels), spq10_labels)) %>%
  select(FID, IID, starts_with("SPQ_"))

write.table(SPQ_pheno, file = here("UKB_SPQ_pheno.txt"), row.names = FALSE, col.names = TRUE, quote=FALSE)
```

```{r discrete COVAR, eval=FALSE, include=FALSE}
#   22001-0.0 Genetic Sex
SPQ_COVAR = SPQ_master %>% select(FID, IID, '22001-0.0') %>% rename(gSEX = '22001-0.0')
write.table(SPQ_COVAR, file = here("UKB_SPQ_COVAR.txt"), row.names = FALSE, col.names = TRUE, quote=FALSE)
```

```{r continous qCOVAR, eval=FALSE, include=FALSE}
# age_years
#  "22000-0.0"   # Genotype batch
#  genetic_pca_vars

SPQ_qCOVAR = SPQ_master %>%
  select(FID, IID, age_years,'22000-0.0', all_of(genetic_pca_vars)) %>%
  rename(gbatch = '22000-0.0') %>%
  rename_with(~ paste0("gPCA", str_extract(., "(?<=\\.)\\d+")), starts_with("22009"))
write.table(SPQ_qCOVAR, file = here("UKB_SPQ_qCOVAR.txt"), row.names = FALSE, col.names = TRUE, quote=FALSE)
```

```{bash generate mbfile, eval=FALSE, include=FALSE}
for i in $(seq 1 22); do
  echo "/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/Bed/chr${i}"
done > /rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/SPQ/SPQ_pre_GWAS/UKB_SPQ_GWAS_mbfile.txt
```

# GWAS: construct and submitting slurm job
```{r construct GWAS slurm code, eval=FALSE, include=FALSE}
source(here("slurm_scheduler.R")) # import slurm scheduler function

out_dir = "/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/SPQ/SPQ_GWAS/"

GWAS_command = paste0(
  "COLNAME=$(awk -v col=\"$SLURM_ARRAY_TASK_ID\" 'NR==1{print $(col+2)}' UKB_SPQ_pheno.txt)", "\n",
  "/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/toolbox/gcta/gcta64 --fastGWA-mlm", " ",
  "--mbfile", " ", "UKB_SPQ_GWAS_mbfile.txt", " ",
  "--grm-sparse", " ", glue("{path_UKB.vw}GRM/UKB_AQ_GRM_SPARSE"), " ",
  "--pheno", " ", "UKB_SPQ_pheno.txt", " ",
  "--mpheno $SLURM_ARRAY_TASK_ID", " ",
  "--qcovar", " ", "UKB_SPQ_qCOVAR.txt", " ",
  "--covar", " ", "UKB_SPQ_COVAR.txt", " ",
  "--out", " ", glue("{out_dir}GWAS_$COLNAME")
)

# construct slurm code:
# Sensitive info are imported through .Rprofile therefore is remained hidden in this code.
GWAS_job_name = "YG_SPQ_GWAS"
time = "8:00:00"
mem = "100G"
home_dir = "/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/SPQ/SPQ_pre_GWAS"
log_dir = "/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/SPQ/SPQ_GWAS"
array = "1-12"

submit_slurm(account = account, 
             home_dir = home_dir, 
             log_dir = log_dir,
             job_name = GWAS_job_name, 
             command = GWAS_command, 
             time = time, 
             mem = mem,
             array = array)

# Then in terminal do this -> sbatch /rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/SPQ/SPQ_GWAS/YG_SPQ_GWAS.sh 
```

# GWAS Manhattan Visualisation
```{R QQ and manhattan plot, eval=FALSE, include=FALSE}
source("/rds/user/yg330/rds-genetics_hpc-Nl99R8pHODQ/UKB/SPQ/SPQ_post_GWAS/manhattan_and_QQ/topr_manhattan_QQ_plot_generate.R")
```